Modeling the effect of aneuploidy on cancer evolution

### 2023-05-31 EvolutionaryRescue1.tex
start merging figures

To estimate the rescue probability $p_{rescue}$, we assume independence between clonal lineages starting from an initial population of $N$ wildtype cells (we check the effect of density-dependent growth on our results below).
Thus, the rescue probability is given by 
\begin{equation}
p_{rescue}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w} ,
\end{equation}
where the approximation $(1-p_w)\approx e^{-p_w}$ assumes that $p_w$ (but not $N p_w$) is small.

Applying the approximations for the survival probability $p_w$ from \cref{survprobwapprox1,survprobwinitial,scenario3} and substituting $\Delta_k=\lambda_k-\mu_k$, we find that the rescue probability can be approximated by
\begin{equation}\label{rescue_prob}
\begin{aligned}
&p_{rescue} \approx \\
  &\begin{cases}
  1-\exp\left[\frac{N}{\Delta_w-u-v}\left(v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right)\right] ,&
  4\lambda_avp_m>\left(\Delta_a-v\right)^2 ,\\
   1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right] ,&
   \Delta_a<0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2 ,\\
   1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right] ,&
   \Delta_a>0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2 .
  \end{cases}
\end{aligned}
\end{equation}

@@

We validate these approximations by comparing them to results of stochastic evolutionary simulations. We find that the approximations work very well (\Cref{rescue_prob_wt_growth_small_N,rescue_prob_wt_growth_large_N,rescue_prob_an_growth_small_N,rescue_prob_an_growth_large_N,rescue_prob_N_nongrow,rescue_prob_N_logistic}).

@@

@@

\caption{Plot of the probability of evolutionary rescue of a population, consisting initially of $N$ wildtype cells, as a function of the proliferation rate of the wildtype cells for various values of the proliferation rate of the aneuploid cells. The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximation \eqref{aneuploidyresistentfirstapprox} for the upper plot and \eqref{AneuploidyDeleteriousApprox} for the lower plot. The dots represent numerical simulations where the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations. Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $N=10^4, \lambda_a=1+10^{-2},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$. }
\label{rescue_prob_wt_growth_small_N}

@@

\caption{Plot of the probability of evolutionary rescue of a population, consisting initially of $N$ wildtype cells, as a function of the proliferation rate of the wildtype cells for various values of the proliferation rate of the aneuploid cells. The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximation \eqref{aneuploidyresistentfirstapprox} for the upper plot and \eqref{AneuploidyDeleteriousApprox} for the lower plot. The dots represent numerical simulations where the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.  Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $N=10^8, \lambda_w=0.14, \lambda_a=0.14,\lambda_m=0.14,\mu_m=0.13, u=10^{-2}, v=10^{-7}$. }
\label{rescue_prob_wt_growth_large_N}

>>>>

To estimate the rescue probability $p_{rescue}$, we assume independence between clonal lineages starting from an initial population of $N$ wildtype cells (we check the effect of density-dependent growth on our results below).
Thus, the rescue probability is given by 
\begin{equation}\label{rescue_prob}
p_{rescue}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w} ,
\end{equation}
where the approximation $(1-p_w)\approx e^{-p_w}$ assumes that $p_w$ (but not $N p_w$) is small.

Applying the approximations for the survival probability $p_w$ from \cref{survprobwapprox1,survprobwinitial,scenario3} and substituting $\Delta_k=\lambda_k-\mu_k$, we find that the rescue probability can be approximated by
\begin{equation}\label{rescue_prob_approx}
\begin{aligned}
&p_{rescue} \approx \\
  &\begin{cases}
  1-\exp\left[\frac{N}{\Delta_w-u-v}\left(v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right)\right] ,&
  4\lambda_avp_m>\left(\Delta_a-v\right)^2 ,\\
   1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right] ,&
   \Delta_a<0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2 ,\\
   1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right] ,&
   \Delta_a>0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2 .
  \end{cases}
\end{aligned}
\end{equation}

@@

We validate these approximations by comparing them to results of stochastic evolutionary simulations. We find that the approximations work very well (\Cref{rescue_prob_wt_growth_small_large_N,rescue_prob_an_growth_small_N,rescue_prob_an_growth_large_N,rescue_prob_N_nongrow,rescue_prob_N_logistic}).

@@

\begin{subfigure}{0.5\textwidth}

@@

\end{subfigure}
\begin{subfigure}{0.5\textwidth}
\end{subfigure}
\caption{
\textbf{Evolutionary rescue probability increases with wildtype and aneuploid growth rates.}
Reduce probability is very high when aneuploidy provides partial resistance, in an initially small tumor (\textbf{Aleft}, $N=10^4$) and even more so in an initially large tumor (\textbf{Aright}, $N=10^8$). % TODO change panel labels to ABCD
When aneuploidy provides tolerance (\textbf{Bleft}, $N=10^4$; (\textbf{Bright}, $N=10^8$), the rescue probability is much lower. However, in all cases rescue probability increase with both the wildtype growth rate (x-axis) and the aneuploidy growth rate (colors).
Markers represent simulation results with 95\% CI; solid and dashed lines for the exact formula (\cref{survival_prob} in \cref{rescue_prob}); % TODO do you use 1-(1-pw)^N or 1-e(-Np_w)?
dashed lines for the approximate formula (\cref{rescue_prob_approx}), demonstrating that they all agree.
% TODO add parameter valuesEvolutionaryRescue1.tex
}
%\caption{Plot of the probability of evolutionary rescue of a population, consisting initially of $N$ wildtype cells, as a function of the proliferation rate of the wildtype cells for various values of the proliferation rate of the aneuploid cells. The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximation \eqref{aneuploidyresistentfirstapprox} for the upper plot and \eqref{AneuploidyDeleteriousApprox} for the lower plot. The dots represent numerical simulations where the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations. Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $N=10^4, \lambda_a=1+10^{-2},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$. }
\label{rescue_prob_wt_growth_small_large_N}

@@

%% large population size
%% A: partial resistance
%% B: tolerance
%\includegraphics[width=1\textwidth]{Figures/CombinedSubplotLargePop.pdf}
%\caption{Plot of the probability of evolutionary rescue of a population, consisting initially of $N$ wildtype cells, as a function of the proliferation rate of the wildtype cells for various values of the proliferation rate of the aneuploid cells. The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximation \eqref{aneuploidyresistentfirstapprox} for the upper plot and \eqref{AneuploidyDeleteriousApprox} for the lower plot. The dots represent numerical simulations where the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.  Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $N=10^8, \lambda_w=0.14, \lambda_a=0.14,\lambda_m=0.14,\mu_m=0.13, u=10^{-2}, v=10^{-7}$. }
%\label{rescue_prob_wt_growth_large_N}
